Collective excitations in competing phases in two and three dimensions

### 2023-12-20 paper/main.tex
uni

@@

\begin{figure*}
    \centering
    \includegraphics[width=.98\textwidth]{plots/resolvent_overview_SC_CDW.pdf}
    \caption{Spectral functions with respect to the operators listed in \eqref{eqn:resolvent_bases}.
    The left column shows the results for the square lattice while the right column shows the results for the simple cubic lattice.
    The left edge of each plot is at $-0.05$ rather than $0$ in order to improve the visibility of the phase peak.
    The shaded area marks the two-particle continuum.
    The panels (a) and (b) show the spectral functions in the SC and CDW phases, respectively.
    The parameters are set to $U=-2.5t$, $V=\pm0.1t$, and $N_\gamma = 3000$.}
    \label{fig:resolvent_overview_SC}
\end{figure*}

In this section, we will be investigating four different diagonal Green's functions $G_{AA^\dagger}(\omega + i0^+)$, namely with
\begin{subequations}
    \label{eqn:resolvent_bases}
    \begin{align}
        A_\text{Higgs} &= \frac{1}{\sqrt{N}} \sum_{\vk} \left( f_{\vk} + f_{\vk}^\dagger \right)  
            = \int \dgamma \left( f_{\gamma} + f_{\gamma}^\dagger \right) \,,\\
        A_\text{Phase} &= \frac{i}{\sqrt{N}} \sum_{\vk} \left( f_{\vk} - f_{\vk}^\dagger \right) = \int \dgamma \left( f_{\gamma} - f_{\gamma}^\dagger \right) \,,\\
        A_\text{CDW}   &= \frac{1}{\sqrt{N}} \sum_{\vk} \left( g_{\vk \up} + g_{\vk \down} \right) = \int \dgamma \left( g_{\gamma \up} + g_{\gamma \down} \right) \,,\\
        A_\text{AFM}   &= \frac{1}{\sqrt{N}} \sum_{\vk} \left( g_{\vk \up} - g_{\vk \down} \right) = \int \dgamma \left( g_{\gamma \up} - g_{\gamma \down} \right) \,,
    \end{align}
\end{subequations}
where $N$ is the number of lattice sites in the system. 
Each one of these operators generates a different kind of collective mode.
The first operator captures the amplitude mode of the $s$-wave superconducting state, while the second one catches the phase mode \cite{Fan22}.
The remaining two capture the collective behavior of the CDW and AFM states, respectively.

@@

\begin{figure*}
    \centering
    \includegraphics[width=.98\textwidth]{plots/resolvent_overview_AFM_CDW.pdf}
    \caption{Spectral functions with respect to the operators listed in \eqref{eqn:resolvent_bases}.
    The left column shows the results for the square lattice while the right column shows the results for the simple cubic lattice.
    The shaded area marks the two-particle continuum.
    The panels (a) and (b) show the spectral functions in the CDW and AFM phases, respectively.
    The parameters are set to $U=zV \pm 0.1t$, $V=1t$, and $N_\gamma = 3000$.}
    \label{fig:resolvent_overview_AFM}
\end{figure*}

To begin, let us describe the spectral functions $\mathcal{A}(\omega) = - (1/\pi) \Im [\mathcal{G} (\omega + i0^+)]$ with respect to the operators above in the various phases.
A plot of them in the SC and CDW phase, close to the phase transition, is shown in \autoref{fig:resolvent_overview_SC}.
We chose $U=-2.5t$ and $V=\pm0.1t$ for both lattices.

There are a variety of different features. The difference between the two lattices is mostly quantitative.
First and foremost, in the SC phase, there is a sharp peak located at $\omega=0$, only in the phase spectral function,
and another peak, only in the amplitude spectral function, located at $\omega=2\Delta_\text{SC}$.
We identify them with the well-known Leggett and Higgs modes in superconductors.
In the CDW phase, both of these spectral functions become identical and form a sharp peak below the two-particle continuum.

The CDW spectral function shows a peak below the continuum only in the SC phase while having another peak at $\omega=2\Delta$ in the CDW phase.
The AFM spectral function is restricted to the continuum for both phases and both lattices.
For the square lattice, however, there is a peak located at the upper edge of the two-particle continuum.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.48\textwidth]{plots/phase_peak.pdf}
    \caption{Log-log plot of the real part of the $A_\text{Phase}$ Green's function in the SC phase at $U=-2.5t$ and $V=-0.1t$ as well as a linear fit to it.
    The upper (lower) panel shows the results for the square (simple cubic) lattice.
    The function behaves as $1/\omega^2$ indicating that the peak in the spectral function is the derivative of a $\delta$ distribution.}
    \label{fig:phase_peak}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=.48\textwidth]{plots/higgs_peak.pdf}
    \caption{Log-log plot of the $A_\text{Higgs}$ spectral function near $\omega = \omega_0 = 2\Delta_\text{SC}$ in the SC phase at $U=-2.5t$ and $V=-0.1t$..
    We additionally plot the difference between the Higgs and the phase spectral functions and fit it, as we interpret the latter as a background within the two-particle continuum.
    The upper (lower) panel shows the results for the square (simple cubic) lattice.
    The spectral function behaves as $1 / \sqrt{\omega - \omega_0}$.}
    \label{fig:higgs_peak}
\end{figure}

@@

In this appendix section, we show fits to the real and imaginary parts of various Green's functions in the different phases.
All fits are linear of the type $y = ax + b$ in double logarithmic plots, i.e., $a$ describes the power-law behavior of the functions.
Firstly, the real part of the $A_\text{Phase}$ Green's function is shown in \autoref{fig:phase_peak}.
It behaves as $1/\omega^2$ indicating that the peak we find in the spectral function at $\omega=0$ corresponds to the derivative of a $\delta$ distribution.

Continuing, we investigate the divergence at $\omega = 2\Delta$ found in the $A_\text{Higgs}$ spectral function within the SC phase.
It behaves as $1/\sqrt{\omega}$ as shown in \autoref{fig:higgs_peak}.

Lastly, we show all the remaining peaks in \autoref{fig:peak_fits}.
The real part of each one of them behaves as $1/\omega$ in close vicinity to the peak position.
This indicates that the peaks are $\delta$ distributions.
The $A_\text{AFM}$ and $A_\text{CDW}$ peaks close the the phase transition are identical when swapped, i.e., $\mathcal{A}_\text{AFM} (\omega)$ in the CDW phase is the same as $\mathcal{A}_\text{CDW} (\omega)$ in the AFM phase.

@@

Numerically, we can compute the matrix elements like this
\begin{equation}
    \label{eqn:numerical_example}
    (A_i | A_j) = \frac{1}{N} \sum_{\vk \vl} h(\gamma_i - \hat{\gamma}(\vk)) h(\gamma_j - \hat{\gamma}(\vl)) (A_{\vk} | A_{\vl})\,.
\end{equation}
Here, all expressions are finite numbers and do not require further treatment.
Mutatis mutandis, one obtains a similar expression for $(A_i | [H, A_j])$.
Additionally, note, that the expressions $(A_{\vk} | A_{\vl})$ and $(A_{\vk} | [H, A_{\vl}])$ can be reduced to sums of terms that are proportional to $\gamma_i$ or $\gamma_j$, respectively.
Therefore, it is not necessary to implement any sort of momentum space computations.
For example, consider $A_{\vk} = f_{\vk}^\dagger$, then
\begin{equation}
    ( f_{\vk}^\dagger | f_{\vl}^\dagger ) = \delta_{\vk, \vl} \langle 1 - n_{\vk \uparrow} - n_{\vk \downarrow} \rangle \eqqcolon \delta_{\vk, \vl} \alpha(\vk)
\end{equation}
holds. Inserting this into \eqref{eqn:numerical_example} yields
\begin{align}
    (A_i | A_j) &= \frac{1}{N} \sum_{\vk} h(\gamma_i - \hat{\gamma}(\vk)) h(\gamma_j - \hat{\gamma}(\vk)) \alpha(\vk)\,.
\end{align}
As stated in the main text, see \eqref{eqn:equal_expecs}, $\alpha(\vk)$ depends only on $\hat{\gamma}(\vk)$, hence we can switch from the momentum sum to a $\gamma$ integration
\begin{align}
    (A_i | A_j) &= \int \rho(\gamma) h(\gamma_i - \gamma) h(\gamma_j - \gamma) \alpha(\gamma) \dgamma \nonumber \\
        &\approx \rho(\gamma_i) h(\gamma_j - \gamma_i) \alpha(\gamma_i) \nonumber \\
        &= \frac{\delta_{ij}}{\Delta \gamma} \rho(\gamma_i) \alpha(\gamma_i)\,.
\end{align}
Since $\Delta \gamma$ is a positive constant factor, it does not impact any of our matrix operations, i.e., we can pull it out to the front of the expressions $\mM \to \tilde{m} (\Delta \gamma)^2$ and $\mN \to \tilde{n} (\Delta \gamma)^2$.
Then, we can rewrite \eqref{eqn:green_derivation} as
\begin{equation}
    G_{AB^\dagger} = - (\Delta \gamma)^2 \vec{b}^\dagger (0) \left[ \tilde{n} \frac{1}{\tilde{m} + z \tilde{n}} \tilde{n} \right] \vec{a}(0)\,.
\end{equation}
To give a practical example, let us compute $A_\text{Higgs}$ from \eqref{eqn:resolvent_bases}.
We order our operator set such that all $f_\gamma^{(\dagger)}$ are upfront. 
This requires the vector elements $\vec{a}$ and $\vec{b}$ to be
\begin{equation}
    a_i = b_i = \begin{cases}
        1 & i < 2 N_\gamma \\
        0 & \text{otherwise}
    \end{cases}\,,
\end{equation}
where $N_\gamma$ is the number of discretization points.
Using this, we can expand the relation for the Green's function
\begin{align}
    G_{AB^\dagger} (z) &= -(\Delta \gamma)^2 \sum_{i=0}^{2 N_\gamma} \sum_{j=0}^{2 N_\gamma} \left[ \tilde{n} \frac{1}{\tilde{m} + z \tilde{n}} \tilde{n} \right]_{ij} \nonumber \\
        &\approx - \int \dgamma \int \dgamma' \left[ \tilde{n} \frac{1}{\tilde{m} + z \tilde{n}} \tilde{n} \right](\gamma, \gamma')\,.
\end{align}
The first line is an approximation of the integral in the second line, as each matrix element corresponds to a specific value for $\gamma$.
Note, that the sum going up to $2N_\gamma$ rather than $N_\gamma$ merely arises from the fact, that we need to integrate over both $f_\gamma$ and $f_\gamma^\dagger$ to obtain $A_\text{Higgs}$.

>>>>

\newcommand{\greens}[1]{\mathcal{G}_\text{#1} (\omega)}
\newcommand{\spectral}[1]{\mathcal{A}_\text{#1}  (\omega)}

@@

\begin{figure*}
    \centering
    \includegraphics[width=.98\textwidth]{plots/resolvent_overview_SC_CDW.pdf}
    \caption{Spectral functions with respect to the operators listed in \eqref{eqn:resolvent_bases}.
    The left column shows the results for the square lattice while the right column shows the results for the simple cubic lattice.
    The left edge of each plot is at $-0.05$ rather than $0$ in order to improve the visibility of the phase peak.
    The shaded area marks the two-particle continuum.
    For each plot, the parameters are set to $U=-2.5t$ and $N_\gamma = 3000$, while $V$ is varied to according to the text in the individual panels.
    The panels show the spectral functions in the (a) pure SC phase, (b) SC/CDW-coexistence, and (c) and (d) pure CDW phase, respectively.}
    \label{fig:resolvent_overview_SC}
\end{figure*}

In this section, we will be investigating four different diagonal Green's functions $G_{AA^\dagger}(\omega + i0^+)$, namely with
\begin{subequations}
    \label{eqn:resolvent_bases}
    \begin{align}
        A_\text{Higgs} &= \frac{1}{\sqrt{N}} \sum_{\vk} \left( f_{\vk} + f_{\vk}^\dagger \right)  
            = \int \dgamma \left( f_{\gamma} + f_{\gamma}^\dagger \right) \,,\\
        A_\text{Phase} &= \frac{i}{\sqrt{N}} \sum_{\vk} \left( f_{\vk} - f_{\vk}^\dagger \right) = \int \dgamma \left( f_{\gamma} - f_{\gamma}^\dagger \right) \,,\\
        A_\text{CDW}   &= \frac{1}{\sqrt{N}} \sum_{\vk} \left( g_{\vk \up} + g_{\vk \down} \right) = \int \dgamma \left( g_{\gamma \up} + g_{\gamma \down} \right) \,,\\
        A_\text{AFM}   &= \frac{1}{\sqrt{N}} \sum_{\vk} \left( g_{\vk \up} - g_{\vk \down} \right) = \int \dgamma \left( g_{\gamma \up} - g_{\gamma \down} \right) \,,
    \end{align}
\end{subequations}
where $N$ is the number of lattice sites in the system. 
Each one of these operators generates a different kind of collective mode.
The first operator captures the amplitude mode of the $s$-wave superconducting state, while the second one catches the phase mode \cite{Fan22}.
The remaining two capture the collective behavior of the CDW and AFM states, respectively.
In the following, we will refer to the Green's functions generated by these operators as $\mathcal{G}_{i}(\omega) = G_{A_i A_i^\dagger}(\omega)$.

@@

\begin{figure*}
    \centering
    \includegraphics[width=.98\textwidth]{plots/resolvent_overview_AFM_CDW.pdf}
    \caption{Spectral functions with respect to the operators listed in \eqref{eqn:resolvent_bases}.
    The left column shows the results for the square lattice while the right column shows the results for the simple cubic lattice.
    The shaded area marks the two-particle continuum.
    For each plot, the parameters are set to $V=1t$ and $N_\gamma = 3000$, while $U$ is varied to according to the text in the individual panels.
    The panels (a) and (c) show the spectral functions in the AFM phase, and the panels (b) and (d) in the CDW phase.}
    \label{fig:resolvent_overview_AFM}
\end{figure*}

To begin, let us describe the spectral functions $\mathcal{A}_i (\omega) = - (1/\pi) \Im [\mathcal{G}_i (\omega + i0^+)]$ with respect to the operators above in the various phases.
\autoref{fig:resolvent_overview_SC} shows them at $U = -2.5t$ in the SC ($V=-0.1t$) and CDW ($V=0.1t$ and $V=0.5t$) phase, as well as in the SC-CDW coexistence at $V=0$.

There are a variety of different features. The difference between the two lattices is mostly quantitative.
First and foremost, in the SC phase, see panel (a), there is a sharp peak located at $\omega=0$ in $\mathcal{A}_\text{Phase}$,
and another peak, only in $\spectral{Higgs}$, located at $\omega=2\Delta_\text{SC}$.
We identify them with the well-known Leggett and Higgs modes in superconductors.
The former can be found at zero energy because our model describes a neutral superfluid wherein there is no coupling of the phase of the order parameter to the electromagnetic potential.
Therefore, this phase may be chosen arbitrarily and changing it requires no energy which ultimately allows a Goldstone mode to exist in the first place. \textcolor{red}{Zitieren!}
In the CDW phase, both of these spectral functions become identical and form a sharp peak below the two-particle continuum.
Following the Landau theory for continuous phase transitions, this is to be expected as the free energy is expanded in terms of the order parameter as
\begin{equation}
    f(\Delta_\text{SC}) \approx r |\Delta_\text{SC}|^2 + \frac{u}{2} |\Delta_\text{SC}|^4\,,
\end{equation}
with $r > 0$ and $u > 0$ if the system is outside of the superconducting phase. \textcolor{Zitieren!}
In this case, the only minimum is at $\Delta_\text{SC} = 0$. 
That means that both possible modes have to overcome the same energy to be excited in the first place. \textcolor{red}{Passt die Beschreibung so?}

At $V=0$, both the SC and the CDW phase coexist.
Here, $\spectral{Higgs}$ and $\spectral{CDW}$ are identical. Both have a peak at $\omega = 0$ and another one at $\omega = 2\Delta$.

$\spectral{CDW}$ shows a peak below the continuum only in the SC phase corresponding to the finite energy necessary to excite this kind of electronic density modulation.
For moderate sized values of $V>0$, i.e., in the CDW phase, see panel (c) at $V=0.1t$, the aforementioned spectral function has a peak at $\omega=2\Delta$.
This peak becomes hard and moves outside the two-particle continuum as $V$ is increased, see panel (d) at $V=0.5t$.

$\spectral{AFM}$ is restricted to the continuum for both phases and both lattices.
For the square lattice, however, there is a peak located at the upper edge of the two-particle continuum.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.48\textwidth]{plots/phase_peak.pdf}
    \caption{Log-log plot of the real part of $\spectral{Phase}$ in the SC phase at $U=-2.5t$ and $V=-0.1t$ as well as a linear fit to it.
    The upper (lower) panel shows the results for the square (simple cubic) lattice.
    The function behaves as $1/\omega^2$ indicating that the peak in the spectral function is the derivative of a $\delta$ distribution.}
    \label{fig:phase_peak}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=.48\textwidth]{plots/higgs_peak.pdf}
    \caption{Log-log plot of $\spectral{}{Higgs}$ near $\omega = \omega_0 = 2\Delta_\text{SC}$ in the SC phase at $U=-2.5t$ and $V=-0.1t$..
    We additionally plot the difference between $\spectral{}{Higgs}$ and $\spectral{Phase}$ and fit it, as we interpret the latter as a background within the two-particle continuum.
    The upper (lower) panel shows the results for the square (simple cubic) lattice.
    The spectral function behaves as $1 / \sqrt{\omega - \omega_0}$.}
    \label{fig:higgs_peak}
\end{figure}

@@

In this appendix section, we show fits to the real and imaginary parts of various Green's functions in the different phases.
All fits are linear of the type $y = ax + b$ in double logarithmic plots, i.e., $a$ describes the power-law behavior of the functions.
Firstly, the real part of $\greens{}{Phase}$ is shown in \autoref{fig:phase_peak}.
It behaves as $1/\omega^2$ indicating that the peak we find in the spectral function at $\omega=0$ corresponds to the derivative of a $\delta$ distribution.

Continuing, we investigate the divergence at $\omega = 2\Delta$ found in $\spectral{Higgs}$ within the SC phase.
It behaves as $1/\sqrt{\omega}$ as shown in \autoref{fig:higgs_peak}.

Lastly, we show all the remaining peaks in \autoref{fig:peak_fits}.
The real part of each one of them behaves as $1/\omega$ in close vicinity to the peak position.
This indicates that the peaks are $\delta$ distributions.
The peaks of $\spectral{AFM}$ and $\spectral{CDW}$ close the the phase transition are identical when swapped, i.e., $\mathcal{A}_\text{AFM} (\omega)$ in the CDW phase is the same as $\mathcal{A}_\text{CDW} (\omega)$ in the AFM phase.

@@

Numerically, we can compute the matrix elements like this
\begin{equation}
    \label{eqn:numerical_example}
    (A_i | A_j) = \frac{1}{N} \sum_{\vk \vl} h(\gamma_i - \hat{\gamma}(\vk)) h(\gamma_j - \hat{\gamma}(\vl)) (A_{\vk} | A_{\vl})\,.
\end{equation}
Here, all expressions are finite numbers and do not require further treatment.
Mutatis mutandis, one obtains a similar expression for $(A_i | [H, A_j])$.
Additionally, note, that the expressions $(A_{\vk} | A_{\vl})$ and $(A_{\vk} | [H, A_{\vl}])$ can be reduced to sums of terms that are proportional to $\gamma_i$ or $\gamma_j$, respectively.
Therefore, it is not necessary to implement any sort of momentum space computations.
For example, consider $A_{\vk} = f_{\vk}^\dagger$, then
\begin{equation}
    ( f_{\vk}^\dagger | f_{\vl}^\dagger ) = \delta_{\vk, \vl} \langle 1 - n_{\vk \uparrow} - n_{\vk \downarrow} \rangle \eqqcolon \delta_{\vk, \vl} \alpha(\vk)
\end{equation}
holds. Inserting this into \eqref{eqn:numerical_example} yields
\begin{align}
    (A_i | A_j) &= \frac{1}{N} \sum_{\vk} h(\gamma_i - \hat{\gamma}(\vk)) h(\gamma_j - \hat{\gamma}(\vk)) \alpha(\vk)\,.
\end{align}
As stated in the main text, see \eqref{eqn:equal_expecs}, $\alpha(\vk)$ depends only on $\hat{\gamma}(\vk)$, hence we can switch from the momentum sum to a $\gamma$ integration
\begin{align}
    (A_i | A_j) &= \int \rho(\gamma) h(\gamma_i - \gamma) h(\gamma_j - \gamma) \alpha(\gamma) \dgamma \nonumber \\
        &\approx \rho(\gamma_i) h(\gamma_j - \gamma_i) \alpha(\gamma_i) \nonumber \\
        &= \frac{\delta_{ij}}{\Delta \gamma} \rho(\gamma_i) \alpha(\gamma_i)\,.
\end{align}
Since $\Delta \gamma$ is a positive constant factor, it does not impact any of our matrix operations, i.e., we can pull it out to the front of the expressions $\mM \to \tilde{m} (\Delta \gamma)^2$ and $\mN \to \tilde{n} (\Delta \gamma)^2$.
Then, we can rewrite \eqref{eqn:green_derivation} as
\begin{equation}
    G_{AB^\dagger} = - (\Delta \gamma)^2 \vec{b}^\dagger (0) \left[ \tilde{n} \frac{1}{\tilde{m} + z \tilde{n}} \tilde{n} \right] \vec{a}(0)\,.
\end{equation}
To give a practical example, let us compute $\greens{Higgs}$ from \eqref{eqn:resolvent_bases}.
We order our operator set such that all $f_\gamma^{(\dagger)}$ are upfront. 
This requires the vector elements $\vec{a}$ and $\vec{b}$ to be
\begin{equation}
    a_i = b_i = \begin{cases}
        1 & i < 2 N_\gamma \\
        0 & \text{otherwise}
    \end{cases}\,,
\end{equation}
where $N_\gamma$ is the number of discretization points.
Using this, we can expand the relation for the Green's function
\begin{align}
    G_{AB^\dagger} (z) &= -(\Delta \gamma)^2 \sum_{i=0}^{2 N_\gamma} \sum_{j=0}^{2 N_\gamma} \left[ \tilde{n} \frac{1}{\tilde{m} + z \tilde{n}} \tilde{n} \right]_{ij} \nonumber \\
        &\approx - \int \dgamma \int \dgamma' \left[ \tilde{n} \frac{1}{\tilde{m} + z \tilde{n}} \tilde{n} \right](\gamma, \gamma')\,.
\end{align}
The first line is an approximation of the integral in the second line, as each matrix element corresponds to a specific value for $\gamma$.
Note, that the sum going up to $2N_\gamma$ rather than $N_\gamma$ merely arises from the fact, that we need to integrate over both $f_\gamma$ and $f_\gamma^\dagger$ to obtain $\greens{Higgs}$.
